Designing Types for R, Empirically

### 2020-05-16 paper/main.tex
minor

\begin{wrapfigure}{r}{6.1cm}
\footnotesize\begin{tabular}{l|l@{}}\hline
\multicolumn{2}{l}{\bf Vectorized data types:}  \\\hline
\k{logical}   & vector of boolean values\\
\k{integer}   & vector of 32 bit integer values\\
\k{double}    & vector of 64 bit floating points\\
\k{complex}   & vector of complex values\\
\k{character} & vector of strings values\\
\k{raw}       & vector of bytes\\
\k{list}      & vector of values of any type\\\hline
\multicolumn{2}{l}{\bf Scalar data types:}\\\hline
\k{NULL}      &  singleton null value\\
\k{S4}        &  instance of a S4 class \\
\k{closure}   &  function with its environment\\
\k{environment}& mapping from symbol to value \\\hline
\multicolumn{2}{l}{\bf Implementation data types:}\\\hline
\multicolumn{2}{l}{\k{special},
\k{builtin},
\k{symbol},
\k{pairlist},
\k{promise}}\\
\multicolumn{2}{l}{
\k{language},
\k{char},
\k{...},
\k{any},
\k{expression},
}\\
\multicolumn{2}{l}{
\k{externalprt},
\k{bytecode},
\k{weakref}}\\\hline
\end{tabular}\caption{Builtin Types}\label{rtypes}\end{wrapfigure}

@@

\code{dplyr::group_indices}: \FUN{\CLASS{\texttt{data.frame}}, \VARGS}{\VEC\int}

\code{lava::index}: \FUN{\CLASS{\texttt{lvm}} \; | \; \CLASS{\texttt{lvmfit}}, \VARGS}{\; ? \; \LIST{\any}}

\code{diptest::dip}: \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl, \dbl}{\CLASS{\texttt{dip}} \; | \; \dbl}

\code{stabledist::cospi2}: \FUN{\VEC\dbl}{\VEC\dbl}

\code{matrixcalc::matrix.power}: \FUN{\CLASS{\texttt{matrix}}, \dbl}{\CLASS{\texttt{matrix}}}

\code{data.tree::Traverse}: \FUN{\CLASS{\texttt{Node}, \texttt{R6}}, ~\VEC\chr, \any, \any}{\LIST{\any}}

\code{openssl::decrypt\_envelope}: \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw, \CLASS{\texttt{key}, \texttt{rsa}}, \any}{\VEC\raw}

\code{dbplyr::set\_win\_current\_group} \FUN{? ~\VEC\chr}{\; ? \; \VEC\chr}

\code{openssl::sha256}: \FUN{~\VEC\raw, ? ~\VEC\raw}{\VEC\raw}

\code{forecast::initparam}: \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl, ~\VEC\dbl, ~\VEC\dbl, \any}{\VEC\dbl}

@@

In the last section, we evaluated our type annotation language based on a
set of function types that were automatically generated by \typetracer.
Now, we're interested in knowing how representative our trace-generated
types are of the functions they were generated for.  To measure this, we
conducted a large-scale experiment: for each package in the corpus discussed
in Section~\ref{sec:corpus}, we ran \typetracer on its test, example, and
vignette code, obtaining types for package functions. Then, we
ran the test, example, and vignette code of each of the package's reverse
dependencies (or clients).  We collected failed contract assertions in order
to analyze them and perhaps identify weaknesses in our type annotation
language.

We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS
total assertions.  Overall, we found that only \PERCFAILEDASSERTIONSWUNDEF
of contract assertions failed.  We made a simplification in our analysis,
capping the number of function arguments we generate types for to 20, and
this resulted in only \PERCASSERTIONSUNDEF failed assertions, which suggests
that the simplification we made did not greatly affect the results.  Once we
controlled for these, we found that \PERCFAILEDASSERTIONS of all assertions
failed for reasons related to our generated types.  Additionally, we found
that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types
were never violated.  The number of immaculate function types increases to
\PERCSUCCFUNNOSTHREE if we discount S3 functions: These S3 functions
represent user-defined polymorphism which we do not tackle in this work.
Overall, these numbers are promising, and suggest that the type signatures
generated by running \typetracer on our corpus are indeed representative,
and capture intended function behaviour.

We breakdown the failed assertions by type in
Table~\ref{tbl:fail-breakdown-0}.  
Accounting for 27.37\% of assertion failures, the most common vilation arises when a value with \CLASS{\texttt{matrix}} is passed where a \VEC\dbl is expected.
Considering these types, we might imagine them to be compatible, however not allowing this coercion was a deliberate design decision:
coercion from vectors to matrices is ad hoc at best, and surely not a
practice codified in the language.  In a similar vein, another popular
failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$,
another case of commonly performed coercion.  We did not include these types
of coercions in our type annotation framework as programmers cannot rely on
them, and it is not always the case that the coercions are safe to perform.

\begin{table}
% BEGIN Autogenerated

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & Occurrences & \% Total & Cumulative \%\\
\midrule
\VEC\dbl & \CLASS{\texttt{matrix}} & 371,896 & 27.37 & 27.37\\
\VEC\dbl & \VEC\int & 117,212 & 8.62 & 35.99\\
\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 100,100 & 7.37 & 43.36\\
\dbl & \int | \null & 60,578 & 4.46 & 47.81\\
\VEC\dbl & \CLASS{\texttt{timeSeries}} & 58,872 & 4.33 & 52.15\\
\addlinespace
\CLASS{\texttt{matrix}} & \CLASS{\texttt{timeSeries}} & 55,005 & 4.05 & 56.19\\
\VEC\dbl & \dbl & 53,338 & 3.92 & 60.12\\
\dbl & \CLASS{\texttt{data.frame}} & 32,553 & 2.40 & 62.51\\
\VEC\dbl & \CLASS{\texttt{data.frame}} & 31,576 & 2.32 & 64.84\\
\dbl & \int & 20,586 & 1.51 & 66.35\\
\bottomrule
\end{tabular}
% END Autogenerated
\label{tbl:fail-breakdown-0}
\caption{Top types of contract assertion failures.}
\end{table}

In addition to the raw number of failing contract checks, we were interested
in how many functions had a parameter where a contract check failed, and
overall we found this to be the case in \PROPFUNSFAILEDCHECK of functions.
Digging a little deeper, we noticed that many of these functions were
performing S3 dispatch, which is part of R's object-orientation framework:
thus, these functions implement user-defined polymorphism which we do not
handle in our type framework.  Removing those functions, we see that
\PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.
These are functions which were under-tested, as calls to these functions
represent only \PERCCALLSBADFUNSINTESTS of recorded calls during
\typetracer's run on the core corpus to infer types.

Turning our attention now to arguments, we found that only
\PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting
runtime occurrences, but that data alone does not tell the full story, as
some failures may be overrepresented if e.g. a failing contract assertion
was in a loop.  We were interested in knowing for each of the most common
violations in Table~\ref{tbl:fail-breakdown-0}, how many different arguments
had that type, and how many of those exhibited the contract failure in
question.  Table~\ref{tbl:fail-breakdown-1} breaks down the failed assertions by type,
folding away multiple identical failed contract assertions for the same parameter
position.  The first row of this table reads: a value of type
\VEC\dbl was passed to 15 different function parameters expecting a
\CLASS{\texttt{matrix}}, of which there are 1394 in total: 1.08\% of \CLASS{\texttt{matrix}}-typed parameters
were passed \VEC\dbl values.

We see that even though the double vector and matrix issue was wildly
prevalent in the raw, dynamic contract evaluation numbers, the number of
actual function argument types that were violated is very small.  The story
is similar with the double and integer coercion we mentioned earlier: it
represents many dynamic contract failures, but very few of the \VEC\int-typed 
arguments have their contracts violated by \VEC\dbl-typed
values.  Rows 5 and 6 are interesting: we see that rather often arguments
expecting \CLASS{\texttt{timeSeries}} data are passed \VEC\dbl or \CLASS{\texttt{matrix}} values.  This
is a quirk of the \code{timeSeries} package, whose functions often accept
matrices and vectors, converting them to time series in an ad hoc manner.
Note that the code coverage of the \code{timeSeries} tests, examples, and
vignettes on the \code{timeSeries} package code is only 58.24\%, which is
one possible explanation of why these contract failures are occurring: the
types that \typetracer generates are only as good as the test code its run
on.

 \begin{table}
% BEGIN Autogenerated

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
\VEC\dbl & \CLASS{\texttt{matrix}} & 15 & 1394 & 1.08\\
\VEC\dbl & \VEC\int & 44 & 657 & 6.70\\
\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 1 & 3 & 33.33\\
\dbl & \int | \null & 5 & 20 & 25.00\\
\VEC\dbl & \CLASS{\texttt{timeSeries}} &19 & 35 & 54.29\\
\addlinespace
\CLASS{\texttt{matrix}} & \CLASS{\texttt{timeSeries}} &  19 & 35 & 54.29\\
\VEC\dbl & \dbl & 100 & 5021 & 1.99\\
\dbl & \CLASS{\texttt{data.frame}} &  1 & 900 & 0.11\\
\VEC\dbl & \CLASS{\texttt{data.frame}} &  6 & 900 & 0.67\\
\dbl & \int & 48 & 529 & 9.07\\
\bottomrule
\end{tabular}

% END Autogenerated
 \label{tbl:fail-breakdown-1}
 \caption{Results from Table~\ref{tbl:fail-breakdown-0}, broken down by occurrences of the expected type as a parameter type.}
 \end{table}

@@

 \begin{table}
% BEGIN Autogenerated

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
\VEC\dbl & \VEC\int & 44 & 657 & 6.70\\
\NAVEC\lgl & \null & 158 & 2756 & 5.73\\
\VEC\chr & \chr & 183 & 4102 & 4.46\\
\chr & \null & 117 & 2756 & 4.25\\
\dbl & \null & 107 & 2756 & 3.88\\
\bottomrule
\end{tabular}
% END Autogenerated
 \label{tbl:fail-breakdown-2}
 \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 90th percentile.}
 \end{table}

  \begin{table}
% BEGIN Autogenerated

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
\CLASS{\texttt{data.frame}} & \CLASS{\texttt{data.frame}, \texttt{tbl}, \texttt{tbl\_df}} & 63 & 284 & 22.18\\
\CLASS{\texttt{matrix}} & \CLASS{\texttt{array}} & 13 & 70 & 18.57\\
\NAVEC\lgl & \lgl | \null & 9 & 49 & 18.37\\
\VEC\chr & \LIST{\VEC\chr} & 20 & 109 & 18.35\\
\VEC\clx & \clx & 7 & 42 & 16.67\\
\bottomrule
\end{tabular}
% END Autogenerated
 \label{tbl:fail-breakdown-3}
 \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile.}
 \end{table}

>>>>

\begin{wrapfigure}{r}{6.1cm}
\footnotesize\begin{tabular}{l@{}l@{}}\hline
\bf Vectors:\\\hline
\k{logical}   & vector of boolean values\\
\k{integer}   & vector of 32 bit integer values\\
\k{double}    & vector of 64 bit floating points\\
\k{complex}   & vector of complex values\\
\k{character} & vector of strings values\\
\k{raw}       & vector of bytes\\
\k{list}      & vector of values of any type\\\hline
{\bf Scalars:}\\\hline
\k{NULL}      &  singleton null value\\
\k{S4}        &  instance of a S4 class \\
\k{closure}   &  function with its environment\\
\k{environment}& mapping from symbol to value \\\hline
{\bf Implementation:}\\\hline
\k{special},
\k{builtin}, &\k{symbol}, \k{pairlist}, \k{promise}\\
\k{language}, \k{char}, \k{...}, & \k{~any}, \k{expression},\\
\k{externalprt},& \k{bytecode}, \k{weakref}\\\hline
\end{tabular}\caption{Builtin Types}\label{rtypes}\end{wrapfigure}

@@

\begin{table}[H]
  \centering
  \resizebox{.8\linewidth}{!}{
  \begin{tabular}{llrrr}
    \toprule
    Function & Type Signature \\
    \midrule
    \code{dplyr::group_indices} & \FUN{\CLASS{\texttt{data.frame}}, \VARGS}{\VEC\int}\\
    \code{moments::all.cumulants} & \FUN{\CLASS{\texttt{matrix}} \; | \; \VEC\dbl}{\CLASS{\texttt{matrix}}\; | \; \VEC{\dbl}}\\
    \code{diptest::dip}& \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl,
                         \dbl}{\CLASS{\texttt{dip}} \; | \; \dbl} \\
    \code{stabledist::cospi2}& \FUN{\VEC\dbl}{\VEC\dbl} \\
    \code{matrixcalc::matrix.power}& \FUN{\CLASS{\texttt{matrix}},
                                     \dbl}{\CLASS{\texttt{matrix}}} \\
    \code{data.tree::Traverse}& \FUN{\CLASS{\texttt{Node}, \texttt{R6}}, ~\VEC\chr,
                                \any, \any}{\LIST{\any}} \\
    \code{openssl::decrypt\_envelope}& \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw,
                                       \CLASS{\texttt{key}, \texttt{rsa}}, \any}{\VEC\raw} \\
    \code{dbplyr::set\_win\_current\_group}& \FUN{? ~\VEC\chr}{\; ? \; \VEC\chr} \\
    \code{openssl::sha256}& \FUN{~\VEC\raw, ? ~\VEC\raw}{\VEC\raw} \\
    \code{forecast::initparam}& \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl,
                                ~\VEC\dbl, ~\VEC\dbl, \any}{\VEC\dbl} \\
    \bottomrule
  \end{tabular}}
\caption{Select Type Signatures}\label{tbl:typesig}
\end{table}

@@

How robust are the inferred types?  To measure this, we conducted another
large-scale experiment: for each package in the corpus, we ran used the
inferred type signatures as contracts and ran all of the CRAN reverse
dependencies for that package. In total we ran \NUMPKGSEVAL packages and
recorded \NUMASSERTIONS total assertions.  Overall, we found that only
\PERCFAILEDASSERTIONSWUNDEF of contract failed.  The limit on number of
arguments (we record only 20) accounted for \PERCASSERTIONSUNDEF failed
assertion. We found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS
of functions types never failed.  The number of immaculate function types
increases to \PERCSUCCFUNNOSTHREE if we discount S3 object method dispatch.
Overall, these numbers are promising, and suggest that the type signatures
are robust.

We breakdown the failed assertions by type in
Table~\ref{tbl:fail-breakdown-0}.  Accounting for 27\% of assertion
failures, is when \CLASS{\texttt{matrix}} is passed where a \VEC\dbl is
expected.  Considering these types, we might imagine them to be compatible,
however not allowing this coercion was a deliberate design decision:
coercion from vectors to matrices is ad hoc at best, and surely not a
practice codified in the language.  In a similar vein, another popular
failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$,
another case of commonly performed coercion.  We did not include these types
of coercions in our type annotation framework as programmers cannot rely on
them, and it is not always the case that the coercions are safe to perform.

\begin{table}
% BEGIN Autogenerated
\small
\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & Occurrences & \% Total & Cumulative \%\\
\midrule
\VEC\dbl & \CLASS{\texttt{matrix}} & 371,896 & 27.37 & 27.37\\
\VEC\dbl & \VEC\int & 117,212 & 8.62 & 35.99\\
\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 100,100 & 7.37 & 43.36\\
\dbl & \int | \null & 60,578 & 4.46 & 47.81\\
\VEC\dbl & \CLASS{\texttt{timeSeries}} & 58,872 & 4.33 & 52.15\\
\addlinespace
\CLASS{\texttt{matrix}} & \CLASS{\texttt{timeSeries}} & 55,005 & 4.05 & 56.19\\
\VEC\dbl & \dbl & 53,338 & 3.92 & 60.12\\
\dbl & \CLASS{\texttt{data.frame}} & 32,553 & 2.40 & 62.51\\
\VEC\dbl & \CLASS{\texttt{data.frame}} & 31,576 & 2.32 & 64.84\\
\dbl & \int & 20,586 & 1.51 & 66.35\\
\bottomrule
\end{tabular}
% END Autogenerated
\label{tbl:fail-breakdown-0}\caption{Top contract failures}\end{table}

In addition to the number of failed contract checks, we were interested in
how many functions had a parameter where a contract check failed, and
overall we found this to be the case in \PROPFUNSFAILEDCHECK of functions.
Digging a little deeper, we noticed that many of these functions were
performing S3 dispatch which we do not handle.  Removing those functions, we
see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing
contract checks.  These are functions were likely under-tested, as calls to
these functions represent only \PERCCALLSBADFUNSINTESTS of recorded calls
during \typetracer's run on the core corpus to infer types.

Turning our attention now to arguments, we found that only
\PROPARGSFAILINGASSERTS of parameters failed.
Table~\ref{tbl:fail-breakdown-0} showed the runtime occurrences, but that
data alone does not tell the full story, as some failures may be
overrepresented if e.g. a failing contract assertion was in a loop.  We were
interested in knowing for each of the most common violations in
Table~\ref{tbl:fail-breakdown-0}, how many different arguments had that
type, and how many of those exhibited the contract failure in question.
Table~\ref{tbl:fail-breakdown-1} breaks down the failed assertions by type,
folding away multiple identical failed contract assertions for the same
parameter position.  The first row of this table reads: a value of type
\VEC\dbl was passed to 15 different function parameters expecting a
\CLASS{\texttt{matrix}}, of which there are 1394 in total: 1\% of
\CLASS{\texttt{matrix}}-typed parameters were passed \VEC\dbl values.

We see that even though the double vector and matrix issue was wildly
prevalent in the raw, dynamic contract evaluation numbers, the number of
actual function argument types that were violated is very small.  The story
is similar with the double and integer coercion we mentioned earlier: it
represents many dynamic contract failures, but very few of the
\VEC\int-typed arguments have their contracts violated by \VEC\dbl-typed
values.  Rows 5 and 6 are interesting: we see that rather often arguments
expecting \CLASS{\texttt{timeSeries}} data are passed \VEC\dbl or
\CLASS{\texttt{matrix}} values.  This is a quirk of the \code{timeSeries}
package, whose functions often accept matrices and vectors, converting them
to time series in an ad hoc manner.  Note that the code coverage of the
\code{timeSeries} tests, examples, and vignettes on the \code{timeSeries}
package code is only 58\%, which is one possible explanation of why these
contract failures are occurring: the types that \typetracer generates are
only as good as the test code its run on.

 \begin{table}
% BEGIN Autogenerated
\small
\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
\VEC\dbl & \CLASS{\texttt{matrix}} & 15 & 1394 & 1.08\\
\VEC\dbl & \VEC\int & 44 & 657 & 6.70\\
\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 1 & 3 & 33.33\\
\dbl & \int | \null & 5 & 20 & 25.00\\
\VEC\dbl & \CLASS{\texttt{timeSeries}} &19 & 35 & 54.29\\
\addlinespace
\CLASS{\texttt{matrix}} & \CLASS{\texttt{timeSeries}} &  19 & 35 & 54.29\\
\VEC\dbl & \dbl & 100 & 5021 & 1.99\\
\dbl & \CLASS{\texttt{data.frame}} &  1 & 900 & 0.11\\
\VEC\dbl & \CLASS{\texttt{data.frame}} &  6 & 900 & 0.67\\
\dbl & \int & 48 & 529 & 9.07\\
\bottomrule
\end{tabular}

% END Autogenerated
 \label{tbl:fail-breakdown-1}
 \caption{Results from Table~\ref{tbl:fail-breakdown-0}, broken down by occurrences of the expected type as a parameter type}
 \end{table}

@@

 \begin{table}
% BEGIN Autogenerated\
\small
\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
\VEC\dbl & \VEC\int & 44 & 657 & 6.70\\
\NAVEC\lgl & \null & 158 & 2756 & 5.73\\
\VEC\chr & \chr & 183 & 4102 & 4.46\\
\chr & \null & 117 & 2756 & 4.25\\
\dbl & \null & 107 & 2756 & 3.88\\
\bottomrule
\end{tabular}
% END Autogenerated
 \label{tbl:fail-breakdown-2}
 \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 90th percentile}
 \end{table}

  \begin{table}
% BEGIN Autogenerated
\small
\begin{tabular}{ll@{}rrr@{}}
\toprule
Passed & Arg Type & \#Args Fail & \#Args w Type & \% Failure\\
\midrule
\CLASS{\texttt{data.frame}} & \CLASS{\texttt{data.frame}, \texttt{tbl}, \texttt{tbl\_df}} & 63 & 284 & 22.18\\
\CLASS{\texttt{matrix}} & \CLASS{\texttt{array}} & 13 & 70 & 18.57\\
\NAVEC\lgl & \lgl | \null & 9 & 49 & 18.37\\
\VEC\chr & \LIST{\VEC\chr} & 20 & 109 & 18.35\\
\VEC\clx & \clx & 7 & 42 & 16.67\\
\bottomrule
\end{tabular}
% END Autogenerated
 \label{tbl:fail-breakdown-3}
 \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile}
 \end{table}
